diffEnrich: An \proglang{R} Package to Compare Functional Enrichment Between Two Experimentally-derived Groups of Genes by Connecting to the KEGG REST API

### 2019-11-22 jss-article-rnw-diffEnrich/article.Rnw
Added sup table 2

%% recommended packages
\usepackage{thumbpdf,lmodern}
\usepackage{textcomp}
\usepackage[section]{placeins}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{adjustbox}
\newcommand\mc[1]{\multicolumn{1}{c}{#1}}

@@

By default, p-values from the Fisher’s Exact test are adjusted for multiple
comparisons with a False Discovery Rate (FDR) (Benjamini:1995), however users
have the option of choosing any type of multiple testing correction supported
by \code{p.adjust}. In addition to the unadjusted p-value and FDR, pathEnrich
will calculate for each KEGG pathway, its fold enrichment defined as the ratio
of number of genes observed from the gene list annotated to the KEGG pathway to
the expected number of genes from the gene list to be annotated to the KEGG
pathway by chance. An example of the first 6 results generated by \code{pathEnrich}
are in Supplementary Table 1. For a detailed description of the variables in this table see Table 3.

<<list1_head, echo = FALSE, eval = FALSE>>=
library(xtable)
tb2 <- head(list1_pe$enrich_table)
xtable(tb2, caption = "Table 2. First 6 results generated by pathEnrich.")
@

@@

\begin{table}[h]
  \centering
  \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
  \tiny
  \begin{adjustbox}{width=\textwidth}
  \begin{tabular}{rllrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
  \hline
95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
  172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
  194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
  212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
  197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
  66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

\bibliography{refs}

>>>>

%% recommended packages
\usepackage{thumbpdf,lmodern}
\usepackage{textcomp}
\usepackage[section]{placeins}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{makecell}
\newcommand\mc[1]{\multicolumn{1}{c}{#1}}

@@

By default, p-values from the Fisher’s Exact test are adjusted for multiple
comparisons with a False Discovery Rate (FDR) (Benjamini:1995), however users
have the option of choosing any type of multiple testing correction supported
by \code{p.adjust}. In addition to the unadjusted p-value and FDR, pathEnrich
will calculate for each KEGG pathway, its fold enrichment defined as the ratio
of number of genes observed from the gene list annotated to the KEGG pathway to
the expected number of genes from the gene list to be annotated to the KEGG
pathway by chance. An example of the first 6 results generated by \code{pathEnrich}
are in Supplementary Table 1. For a detailed description of the variables in this table see Supplementary Table 2.

@@

<<list1_head, echo = FALSE, eval = FALSE>>=
library(xtable)
suptb1 <- head(list1_pe$enrich_table)
xtable(suptb1)
@

\begin{table}[ht]
  \centering
  \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
  \tiny
  \begin{adjustbox}{width=\textwidth}
  \begin{tabular}{rllrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
  \hline
95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
  172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
  194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
  212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
  197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
  66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

<<pathEnrich_description, echo=FALSE, eval = FALSE>>=
# get column names
names <- colnames(list1_pe$enrich_table)
# Define column names
description <- c("KEGG Pathway Identifier",
                 "Description of KEGG Pathway (provided by KEGG)",
                 "Number of Genes in KEGG Pathway",
                 "Number of Genes from gene list in KEGG Pathway",
                 "Number of Genes in KEGG Database",
                 "Number of Genes from gene list in KEGG Database",
                 "Expected number of genes from list to be in KEGG pathway by chance",
                 "P-value for enrichment within the KEGG pathway for list genes",
                 "Multiple testing adjusted enrichment p-values (default = False Discovery Rate (Benjamini and Hochberg, 1995))",
                 "Ratio of number of genes observed from the gene list annotated to the KEGG pathway to the number of genes expected from the gene list to be annotated to the KEGG pathway if there was no enrichment (i.e. KEGG_PATHWAY_in_list/expected)")
# Make dataframe
suptb2 <- data.frame(names, description)
colnames(suptb2) <- c("Column Names", "Column Description")

## Generate xtable
xtable(suptb2)
@

\begin{table}[h]
\centering
  \caption{Supplementary Table 2. Description of columns is dataframe generated by pathEnrich.}
  \begin{adjustbox}{width=\textwidth}
  \begin{tabular}{rll}
  \hline
& Column Names & Column Description \\
  \hline
& KEGG\_PATHWAY\_ID & \makecell{KEGG Pathway Identifier \\} \\
& & \\
& KEGG\_PATHWAY\_description & \makecell{Description of KEGG Pathway \\ (provided by KEGG) \\ } \\
  & & \\
& KEGG\_PATHWAY\_cnt & \makecell{Number of Genes in KEGG Pathway} \\
& & \\
& KEGG\_PATHWAY\_in\_list & \makecell{Number of Genes from  \\ gene list in KEGG Pathway} \\
& & \\
& KEGG\_DATABASE\_cnt & \makecell{Number of Genes in KEGG Database} \\
& & \\
& KEG\_DATABASE\_in\_list & \makecell{Number of Genes from \\ gene list in KEGG Database} \\
& & \\
& expected & \makecell{Expected number of genes \\ from list to be in \\ KEGG pathway by chance} \\
& & \\
& enrich\_p & \makecell{P-value for enrichment within \\ the KEGG pathway for list genes} \\
& & \\
& p\_adj & \makecell{Multiple testing adjusted enrichment \\ p-values (default = False Discovery Rate \\ (Benjamini and Hochberg, 1995))} \\
& & \\
& fold\_enrichment & \makecell{Ratio of number of genes observed  \\ from the gene list annotated to the \\ KEGG pathway to the number of genes  \\ expected from the gene list to  \\ be annotated to the KEGG pathway if  \\ there was no enrichment \\ (i.e. KEGG\_PATHWAY\_in\_list/expected)} \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

%% ----------------------------------------------------------------------
\bibliography{refs}

### 2019-11-22 jss-article-rnw-diffEnrich/article.tex
Added sup table 2

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Kegg Release: Release\_92.0+\_11-21\_Nov\_19}}\end{kframe}
\end{knitrout}

@@

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Kegg Release: Release\_92.0+\_11-21\_Nov\_19}}\end{kframe}
\end{knitrout}

@@

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# ncbi\_to\_kegg2019-11-21Release\_92.0+\_11-21\_Nov\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# kegg\_to\_pathway2019-11-21Release\_92.0+\_11-21\_Nov\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# pathway\_to\_species2019-11-21Release\_92.0+\_11-21\_Nov\_19.txt}}

@@

By default, p-values from the Fisher’s Exact test are adjusted for multiple
comparisons with a False Discovery Rate (FDR) (Benjamini:1995), however users
have the option of choosing any type of multiple testing correction supported
by \code{p.adjust}. In addition to the unadjusted p-value and FDR, pathEnrich
will calculate for each KEGG pathway, its fold enrichment defined as the ratio
of number of genes observed from the gene list annotated to the KEGG pathway to
the expected number of genes from the gene list to be annotated to the KEGG
pathway by chance. An example of the first 6 results generated by \code{pathEnrich}
are in Supplementary Table 1. For a detailed description of the variables in this table see Table 3.

@@

\begin{table}[h]
  \centering
  \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
  \tiny
  \begin{adjustbox}{width=\textwidth}
  \begin{tabular}{rllrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
  \hline
95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
  172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
  194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
  212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
  197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
  66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

\bibliography{refs}

>>>>

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# 3 data sets will be written as tab delimited text files}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# File location: /Users/harry/Documents/Saba\_Lab/diffEnrich}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Kegg Release: Release\_92.0+\_11-22\_Nov\_19}}\end{kframe}
\end{knitrout}

@@

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Kegg Release: Release\_92.0+\_11-22\_Nov\_19}}\end{kframe}
\end{knitrout}

@@

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# ncbi\_to\_kegg2019-11-21Release\_92.0+\_11-22\_Nov\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# kegg\_to\_pathway2019-11-21Release\_92.0+\_11-22\_Nov\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# pathway\_to\_species2019-11-21Release\_92.0+\_11-22\_Nov\_19.txt}}

@@

By default, p-values from the Fisher’s Exact test are adjusted for multiple
comparisons with a False Discovery Rate (FDR) (Benjamini:1995), however users
have the option of choosing any type of multiple testing correction supported
by \code{p.adjust}. In addition to the unadjusted p-value and FDR, pathEnrich
will calculate for each KEGG pathway, its fold enrichment defined as the ratio
of number of genes observed from the gene list annotated to the KEGG pathway to
the expected number of genes from the gene list to be annotated to the KEGG
pathway by chance. An example of the first 6 results generated by \code{pathEnrich}
are in Supplementary Table 1. For a detailed description of the variables in this table see Supplementary Table 2.

@@

\begin{table}[ht]
  \centering
  \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
  \tiny
  \begin{adjustbox}{width=\textwidth}
  \begin{tabular}{rllrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
  \hline
95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
  172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
  194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
  212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
  197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
  66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

\begin{table}[h]
\centering
  \caption{Supplementary Table 2. Description of columns is dataframe generated by pathEnrich.}
  \begin{adjustbox}{width=\textwidth}
  \begin{tabular}{rll}
  \hline
& Column Names & Column Description \\
  \hline
& KEGG\_PATHWAY\_ID & \makecell{KEGG Pathway Identifier \\} \\
& & \\
& KEGG\_PATHWAY\_description & \makecell{Description of KEGG Pathway \\ (provided by KEGG) \\ } \\
  & & \\
& KEGG\_PATHWAY\_cnt & \makecell{Number of Genes in KEGG Pathway} \\
& & \\
& KEGG\_PATHWAY\_in\_list & \makecell{Number of Genes from  \\ gene list in KEGG Pathway} \\
& & \\
& KEGG\_DATABASE\_cnt & \makecell{Number of Genes in KEGG Database} \\
& & \\
& KEG\_DATABASE\_in\_list & \makecell{Number of Genes from \\ gene list in KEGG Database} \\
& & \\
& expected & \makecell{Expected number of genes \\ from list to be in \\ KEGG pathway by chance} \\
& & \\
& enrich\_p & \makecell{P-value for enrichment within \\ the KEGG pathway for list genes} \\
& & \\
& p\_adj & \makecell{Multiple testing adjusted enrichment \\ p-values (default = False Discovery Rate \\ (Benjamini and Hochberg, 1995))} \\
& & \\
& fold\_enrichment & \makecell{Ratio of number of genes observed  \\ from the gene list annotated to the \\ KEGG pathway to the number of genes  \\ expected from the gene list to  \\ be annotated to the KEGG pathway if  \\ there was no enrichment \\ (i.e. KEGG\_PATHWAY\_in\_list/expected)} \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

%% ----------------------------------------------------------------------
\bibliography{refs}
